Modeling the Java Card API (or Interface?) in B

### 2012-03-27 papers/B2012-work/javacardAPIpaper.tex

\documentclass{entcs} 
\usepackage{entcsmacro}
\usepackage{graphicx}
\sloppy
% The following is enclosed to allow easy detection of differences in
% ascii coding.
% Upper-case    A B C D E F G H I J K L M N O P Q R S T U V W X Y Z
% Lower-case    a b c d e f g h i j k l m n o p q r s t u v w x y z
% Digits        0 1 2 3 4 5 6 7 8 9
% Exclamation   !           Double quote "          Hash (number) #
% Dollar        $           Percent      %          Ampersand     &
% Acute accent  '           Left paren   (          Right paren   )
% Asterisk      *           Plus         +          Comma         ,
% Minus         -           Point        .          Solidus       /
% Colon         :           Semicolon    ;          Less than     <
% Equals        =3D           Greater than >          Question mark ?
% At            @           Left bracket [          Backslash     \
% Right bracket ]           Circumflex   ^          Underscore    _
% Grave accent  `           Left brace   {          Vertical bar  |
% Right brace   }           Tilde        ~

@@

  \title{Modeling the Java Card API (or Interface?) in B} \author[ufrn]{A. Martins Moreira\thanksref{ALL}\thanksref{anamaria}}
  \author[ufrn]{S. de Oliveira Santos \thanksref{simone}}
  \author[ifrn]{B. Emerson Gurgel Gomes \thanksref{bruno}}
  \author[ufrn]{D. Boris Paul D\'{e}harbe \thanksref{david}}
  \address[ufrn]{DIMAp-UFRN, Natal, RN, Brazil} \thanks[ALL]{INES-CAPES-projeto GD?... o que mais? REUNI?}  \address[ifrn]{IFRN, Currais Novos, RN, Brazil} 
 \thanks[anamaria]{Email:
    \href{mailto:anamaria@dimap.ufrn.br} {\texttt{\normalshape
        anamaria@dimap.ufrn.br}}} 
      \thanks[simone]{Email:
    \href{mailto:simone82@ppgsc.ufrn.br} {\texttt{\normalshape
        simone82@ppgsc.ufrn.br}}}
      \thanks[bruno]{Email:
    \href{mailto:bruno.gurgel@ifrn.edu.br} {\texttt{\normalshape
        bruno.gurgel@ifrn.edu.br}}}
       \thanks[david]{Email:
    \href{mailto:david@dimap.ufrn.br} {\texttt{\normalshape
        david@dimap.ufrn.br}}}

% AMM: Precisa ser arrumado: The ENTCS macro package provides two alternatives to listing authors
% names and addresses.  These are described in detail in the file
% \texttt{instraut.dvi}. Basically, listing each author and his or her
% address in turn, is the simplest method.  But, if there are several
% authors and two or more share the same address (but not all authors
% are at this address), then the method of listing authors first, and
% then the addresses, and of referencing addresses to authors should be
% used.

\begin{abstract} 
  The B method development process starts from an abstract model of a
  set of services that are to be developed, and includes formally
  verified refinements, which introduce implementation details. At a
  final refinement step these implementation details depend on the
  language for which code is to be generated. To rigorously develop
  Java Card Applications with the B method, then, it is necessary to
  model in B the constructs of Java Card, so that the gap between the
  B formally verified implementation model and the Java Card generated
  code is as small as it can be. In the context of the BSmart project,
  we have developed the specification of the Java Card Application
  Programming Interface (Java Card API), to be used in the development
  of Java Card applications with B.
\end{abstract}
\begin{keyword}
  Java Card, smart cards, B method, formal methods
\end{keyword}
\end{frontmatter}
\section{Introduction}\label{intro}

In the context of the BSmart project, we are interested in the use of
B to develop Java Card~\cite{chen:2000} applications.  Java Card is a
restricted and optimized version of Java which allows memory and
processor constrained devices, such as smart cards, to store and run
small applications.  Java Card applications have characteristics
(small size and need for security) which make then good candidates for
formal development with the B method. They also present some
regularity of structure that induced our group into the definition of
a method and a tool which specialize B for development of such
applications, providing greater support for the developer than pure B
tools \cite{atelierB,proB}. They are the BSmart method and tool
presented in \cite{Gomes10}.
%colocar o de 2008 também

Previous works have also dealt with B for Java Card, 
\cite{BOM,JBtools,e cia}, but none has, at our knowledge, specified Java Card
in B so that the formal development includes Java Card aspects. In the
first versions of BSmart \cite{Gomes10}, these aspects were included by the code
ganerator, and in JBTools \cite{JBtools}, the generated code needed to be edited by
the developer to include these aspects.

These works lacked then the ability to formally verify the Java Card
dependent aspects of the application. To be able to change this, a B
specification of these aspects is needed, and this contribution is
what we present in this paper: the B specification of the Java Card
API, with...  This specification can be used with the BSmart tool,
independently, in any B tool, or even as a complement to the official
Java Card documentation \cite{oracle}.

@@

Among the different components of the Java Card platform, the Java Card API is the one without which no refinement into Java Card code could be done.
Suppose, for instance, that we had the B abstract machine of fig \ref{maquina}, which models ...  Of course, this is a very simple example, but it is enough to show the importance of having the model of the Java Card API modeled in B.  As all data in it is implementable in C, for instance, and it is a deterministic specification, B tools can directly generate code for it in C, and the resulting code would be very similar to the original B machine. However,  Java Card code corresponding to it would look like the one presented in fig. \ref{impl java card}. In a rigorous development process, we don't want the code generator to carry out all these non-verified transformations in the code. So, we need to be able to refine the original B machine into something like the B refinement of fig. \ref{B ref usando a API}. Then, code generation would be straightforward, as it should be. This is then the reason why it needed to be the first Java Card component to be modeled in B. (AMM: maybe also cite related work ? WHich are the formalization works that did not focus on the API?)

The Java Card API, in its version 2.2.2, is defined in 14 packages, containing 93 classes and interfaces, as shown in table \ref{incluir a tabela da dissert da simone}. Their interfaces have all been modeled in B, and the resulting models are available at LINK. Differently from  JML and OCL, though, B is not object oriented, and some modeling solutions had to be found to cope with this lack of resources.  We present in the following the main characteristics and restrictions of our models:

>>>>

\documentclass{entcs} 
\usepackage{entcsmacro}
\usepackage{graphicx}

\usepackage{listings}
% Define B sintaxe
\lstdefinelanguage{B}  {
keywords={MACHINE, REFINEMENT, REFINES, IMPLEMENTATION, SEES, IMPORTS, SETS, INCLUDES, EXTENDS, CONSTANTS, CONCRETE_CONSTANTS, CONCRETE_VARIABLES, PROPERTIES, DEFINITIONS, VARIABLES, ABSTRACT_VARIABLES, INITIALISATION, INVARIANT, VALUES, OPERATIONS, BEGIN, PRE, THEN, END, NAT, NAT1, BOOL, MAXINT, IF, ELSE, ELSIF, ANY, WHERE, CHOICE, OR, EITHER, SELECT, VAR, IN},
sensitive=true,% 
alsoletter={\$},% 
comment=[l]{\#},% 
string=[b]",% 
string=[b]'% 
}

% Set listing styles
\lstset{ %
%language=Pseudo,                % the language of the code
basicstyle=\footnotesize\ttfamily,            % the size of the fonts that are used for the code
extendedchars=true,
numbers=left,                   % where to put the line-numbers
numberstyle=\tiny,      % the size of the fonts that are used for the line-numbers 
%stepnumber=2,                   % the step between two line-numbers. If it's 1, each line will be numbered
%numbersep=5pt,                  % how far the line-numbers are from the code
%backgroundcolor=\color{white},  % choose the background color. You must add \usepackage{color}
%showspaces=false,               % show spaces adding particular underscores
%showstringspaces=false,         % underline spaces within strings
%showtabs=false,                 % show tabs within strings adding particular underscores
frame=b,   %trBL,                   % adds a frame around the code
stringstyle=\color{white}\ttfamily,
%linewidth=0.8\textwidth, % width of the line
tabsize=2,                      % sets default tabsize to 2 spaces
columns=fixed,
basewidth=.6em,
xleftmargin=20pt,
xrightmargin=5pt,
%captionpos=b,                   % sets the caption-position to bottom
breaklines=true,                % sets automatic line breaking
showspaces=false,           % Leerzeichen anzeigen ?
showtabs=false,             % Tabs anzeigen ?
%backgroundcolor=\color{lightgray},
keywordstyle=\color{red},
showstringspaces=false
%breakatwhitespace=false,        % sets if automatic breaks should only happen at whitespace
%title=\lstname,                 % show the filename of files included with \lstinputlisting; also try caption instead of title
%escapeinside={\%*}{*)},         % if you want to add a comment within your code
%keywords={}            % if you want to add more keywords to the set
}

\sloppy
% The following is enclosed to allow easy detection of differences in
% ascii coding.
% Upper-case    A B C D E F G H I J K L M N O P Q R S T U V W X Y Z
% Lower-case    a b c d e f g h i j k l m n o p q r s t u v w x y z
% Digits        0 1 2 3 4 5 6 7 8 9
% Exclamation   !           Double quote "          Hash (number) #
% Dollar        $           Percent      %          Ampersand     &
% Acute accent  '           Left paren   (          Right paren   )
% Asterisk      *           Plus         +          Comma         ,
% Minus         -           Point        .          Solidus       /
% Colon         :           Semicolon    ;          Less than     <
% Equals        =3D           Greater than >          Question mark ?
% At            @           Left bracket [          Backslash     \
% Right bracket ]           Circumflex   ^          Underscore    _
% Grave accent  `           Left brace   {          Vertical bar  |
% Right brace   }           Tilde        ~

@@

  \title{Modeling the Java Card API (or Interface?) in B}
  \author[ufrn]{A. Martins Moreira\thanksref{ALL}\thanksref{anamaria}}
  \author[ufrn]{S. de Oliveira Santos \thanksref{simone}}
  \author[ifrn]{B. Gurgel Gomes \thanksref{bruno}} \author[ufrn]{D.
    D\'{e}harbe \thanksref{david}} \address[ufrn]{DIMAp-UFRN, Natal,
    RN, Brazil} \thanks[ALL]{This work is partly supported by CAPES
    and CNPq grants 306033/2009-7, 560014/2010-4 and 573964/2008-4
    (National Institute of Science and Technology for Software
    Engineering---INES, \texttt{www.ines.org.br}).}
\address[ifrn]{IFRN, Currais Novos, RN, Brazil}
\thanks[anamaria]{Email: \href{mailto:anamaria@dimap.ufrn.br}
  {\texttt{\normalshape anamaria@dimap.ufrn.br}}}
\thanks[simone]{Email: \href{mailto:simone82@ppgsc.ufrn.br}
  {\texttt{\normalshape simone82@ppgsc.ufrn.br}}}
\thanks[bruno]{Email: \href{mailto:bruno.gurgel@ifrn.edu.br}
  {\texttt{\normalshape bruno.gurgel@ifrn.edu.br}}}
\thanks[david]{Email: \href{mailto:david@dimap.ufrn.br}
  {\texttt{\normalshape david@dimap.ufrn.br}}}

\begin{abstract} 
  The B method development process starts from an abstract model of a
  set of services that are to be developed, and proceeds with formally
  verified refinements, which introduce implementation details. At a
  final refinement step these implementation details depend on the
  language for which code is to be generated. To rigorously develop
  Java Card Applications with the B method, then, it is necessary to
  model in B the constructs of Java Card, so that the gap between the
  B formally verified implementation model and the Java Card generated
  code be as narrow as possible. In the context of the BSmart project,
  we have developed the specification of the Java Card Application
  Programming Interface (Java Card API), to be used in the development
  of Java Card applications with B.
\end{abstract}
\begin{keyword}
  Java Card, smart cards, B method, formal methods
\end{keyword}
\end{frontmatter}
\section{Introduction}\label{intro}

In the context of the BSmart project, we are interested in the use of
B to develop Java Card~\cite{chen:2000} applications.  Java Card is a
restricted and optimized version of Java which allows memory and
processor constrained devices, such as smart cards, to store and run
small applications.  Java Card applications have characteristics
(small size and need for security) which make them good candidates for
formal development with the B method. They also present some
regularity of structure that led our group to define a method and
implement a tool which specialize B for development of such
applications, providing greater support for the developer than pure B
tools \cite{atelierB,proB}. They are the BSmart method and tool,
presented in \cite{BSmart08,Gomes10}.

Previous works have also dealt with B for Smart Cards,
\cite{Bert:2003,JBtools,tatibouet:lncs,requetBtoC}, but none has, at
our knowledge, specified Java Card in B so that the formal development
includes Java Card aspects. In the first versions of BSmart
\cite{Gomes10}, these aspects were included by the code generator, and
in JBTools \cite{JBtools}, the generated code needed to be edited by
the developer to include these aspects.

These works lacked then the ability to formally verify the Java Card
dependent aspects of the application. To be able to change this, a B
specification of these aspects is needed, and this is the contribution
of this paper: the B specification of the Java Card API, with its
primitive types and all the classes of the 14 packages composing the
API. This specification can be used with the BSmart tool, 
independently, in any B tool, or even as a complement to the official
Java Card documentation \cite{oracle}.

@@

Among the different components of the Java Card platform, the Java
Card API is the one without which no refinement into Java Card code
could be done.  Suppose, for instance, that we had the B abstract
machine of Fig.~\ref{lis:buscard}, which models a bus card with a
variable that keeps track of the current credit the user has (\texttt{balance}) and an operation to add credits to this
balance (\texttt{add\_credit}!!! verificar). This simple example provides enough evidence of the
importance of having a B model of the Java Card API.  As all data in
it is implementable in C, for instance, and it is a deterministic
specification, B tools can directly generate code for it in C, and the
resulting code would be very similar to the original B machine.

\lstset{language=B, caption={B machine Buscard},label=lis:buscard}
\begin{lstlisting}
MACHINE
	Buscard
CONSTANTS
	maximum
PROPERTIES
	maximum : INTEGER & maximum = 100
VARIABLES
	balance
INVARIANT
	balance : INTEGER & balance <= maximum
INITIALISATION
	balance := 0
OPERATIONS
	add(credit) =
	PRE
		credit : INTEGER & credit > 0  & balance + credit <= maximum
	THEN
		balance := balance + credit
	END
END
\end{lstlisting}

However, Java Card code corresponding to it would look like the one
presented in Fig.~\ref{lis:add}. In a rigorous development
process, we do not want the code generator to carry out all these
non-verified transformations in the code. So, we need to be able to
refine the original B machine into something such as the B refinement
of Fig.~\ref{B ref usando a API}. Then, code generation would be
straightforward (as it should be). This is then the reason why it
needed to be the first Java Card component to be modeled in B. (AMM:
maybe also cite related work ? Which are the formalization works that
did not focus on the API?)

\lstset{language=Java, caption={Method add in applet Java Card},label=lis:add}
\begin{lstlisting}
public void add(APDU apdu) {
	byte[] buffer = apdu.getBuffer();
	byte credit = (byte) Util.makeShort((byte) buffer[ISO7816.OFFSET_P1], (byte)buffer[ISO7816.OFFSET_P2]);
	if (!(credit >= 0)) {
		ISOException.throwIt(EXCEPTIONS.negative_credit);
	}
	else if (!(balance + credit <= MAXIMUM)) {
		ISOException.throwIt(EXCEPTIONS.balance_exceeded);
	}
	else {
		balance = (balance + credit);
	}
}
\end{lstlisting}

[[[ PARA SIMONE -> tambem precisamos com urgencia daquela anelise sobre o que muda da 2.2.2 para a 3 ]]]

The Java Card API, in its version 2.2.2, is composed of 14 packages,
containing 93 classes and interfaces, as shown in
Table~\ref{tab:packages}. Their interfaces have all been modeled in B,
and the resulting models are available at the KitSmart project
page\footnote{https://code.google.com/p/kitsmart/}. Differently from
JML and OCL, though, B is not object oriented, and some modeling
solutions had to be found to cope with this lack of resources.  We
present in the following the main characteristics and restrictions of
our models:

\begin{table}[h]
\begin{center}
\caption{Packages of the Java Card API (2.2.2) and the number of classes in each package.}
\begin{tabular}{|l|c||l|c|}
 \hline

\textbf{\small{Package}} & \textbf{\small{classes}} & \textbf{\small{Package}} & \textbf{\small{classes}}\\ \hline \hline
\small{java.io} & \small{1} & \small{javacardx.biometry} & \small{5} \\ \hline
\small{java.lang} & \small{12} & \small{javacardx.crypto }& \small{2} \\ \hline
\small{java.rmi} & \small{2} & \small{javacardx.external} & \small{3} \\ \hline
\small{javacard.framework} & \small{19} & \small{javacardx.framework.math} & \small{3} \\ \hline
\small{javacard.framework.service} & \small{8} & \small{javacardx.framework.tlv} & \small{7} \\ \hline
\small{javacard.security} & \small{27} & \small{javacardx.framework.util} & \small{2} \\ \hline
\small{javacardx.apdu} & \small{1} & \small{javacardx.framework.util.intx} & \small{1} \\

 \hline
\end{tabular}
\label{tab:packages}
\end{center}
\end{table}
